Evaluating contact tracing and isolation to control pandemic potential influenza outbreaks

### 2025-05-01 inst/paper/main.tex
update intro and add references

The likelihood of an outbreak being controlled, in other words going locally-extinct or -eliminated, is in part based on the epidemiological characteristics of the pathogen, and in part on the inventions implemented, both pharmaceutical and non-pharmaceutical. The primary epidemiological determinant for the spread of an infectious disease is the reproduction number, defined as the average number of secondary cases  per primary case. The higher the reproduction number the less likely a disease, without intervention, is to cease transmission in a population. If a pathogen is subcritical (i.e. $R < 1$) it will eventually stop spreading but can still cause transient outbreaks \citep{farringtonDistributionTimeExtinction1999}. Individual-level heterogeneity in transmission also influences the possibility of a pathogen going extinct from stochasticity in the transmission process \citep{lloyd-smithSuperspreadingEffectIndividual2005}. Higher individual-level variation for a given reproduction leading to, on average, more outbreaks being controlled (however a small number of outbreaks may become large due to superspreading). In addition, the likelihood of containment is reduced (assuming $R > 1$) if there are multiple introductions into a populations \citep{kucharskiEarlyDynamicsTransmission2020}. \\

A range of interventions can be deployed to reduce or prevent the transmission of infectious diseases, broadly classified into pharmaceutical and non-pharmaceutical interventions (NPIs). Early in an outbreak -- especially in the case of a novel pathogen -- NPIs are often the first line of defence, as pharmaceutical options may not yet be available. These interventions vary in scope and specificity, from targeted measures such as contact tracing and isolation, to broader actions like school closures, bans on gatherings, and regional or national lockdowns (ref). The success of such interventions in containing an outbreak before it becomes sustained and widespread depends on several factors, including the type of intervention, the delay before implementation, and the intervention’s effectiveness (Longini et al., 2005). For highly transmissible pathogens (e.g. $R>3$) with short generation times (e.g. $<3$ days), swift and effective interventions are required to achieve containment. \\

Contact tracing, the process of finding the contacts of infected individuals is an often used NPI to try and curtail the growth of an epidemic. It can involve forward contact tracing, whereby infected individuals are asked who they have in contact with so their contacts can be followed-up and isolated or quarantined is needed; or backward contact tracing whereby infected individuals are asked who they have been in contact with prior to being infected to identify the source of infection and potentially forward trace from the source. Contact tracing and isolation is a targeted approach, like ring vaccination, that looks to prevent the onward transmission of an infectious disease by tackling infected individuals and their contacts \citep{kucharskiEffectivenessRingVaccination2016} (Whittaker et al., 2024) (ref). Isolation of traced individuals has proved widely effective in controlling different outbreaks with a variety of pathogens differing in their aetiological and epidemiological characteristics (SARS ref; other refs). However, it's important to remember that it's not a panacea for all outbreak scenarios. In epidemics or pandemics where the proportion of cases or contacts missed is high, either because surveillance is patchy or because the number of contacts exceeds the capacity of the system, the efficacy of contact tracing to contain an outbreak diminishes, or at worst, fails completely \citep{dhillonWhenContactTracing2018} (ref on limited capacity of tracing). \\

Outbreaks have been shown to be more easily controlled when a large proportion of onward transmission takes place after the onset of symptoms \citep{fraserFactorsThatMake2004}.

It has become widely accepted that influenza viruses cannot be controlled with contact tracing and isolation become the rate at which infected individuals become infectious and show symptoms is faster than the latency in the process of contacting contacts (ref). However, with the recent human infections of avian influenza (A/H5N1), mostly among poultry and cattle farmers (Garg et al., 2025) and the longer estimates of incubation period and serial interval (SI) from historical H5N1 data \citep{Ward2024.12.11.24318702} it may not be the case that contact tracing and isolation is unable to control influenza. Historically, avian influenza (H5N1 and H7N9) have had high severity (case fatality risk $\sim 35\%-55\%$), and therefore sustained transmission in a human population could have substantial mortality and morbidity implications \citep{tannerPandemicPotentialAvian2015}. \\

The (antigenic) novelty of a pathogen introduced into a human population is a leading risk factor for causing a pandemic (ref). Thus, influenza subtypes that have negligible levels of existing immunity in the population are probable epidemic and pandemic candidates. This is the case for H5N1 and H7N9 \citep{tannerPandemicPotentialAvian2015} and was the case when H1N1 caused an outbreak in 2009 (ref). \\

In this study we evaluate the effectiveness of controlling infectious disease outbreaks that exhibit influenza-like epidemiological characteristics using a mathematical model. In effect the ability to control an epidemic with contact tracing is a race between the identification and isolation of contacts and the spread of the infectious agent (Reyna-Lara et al., 2021). The dogma that influenza contagion outpaces our ability to trace and isolate in time, may not hold for all pathogen subtypes with new estimates of incubation period and serial interval (Ward et al., 2024). This study looks to define the speed and effectiveness of isolation given an epidemic scenario to inform outbreak response complementing in combination with other quickly available measures for pandemic potential influenza strains, such as rapid testing and antivirals (ref).

@@

The NPI implemented in the transmission model is an isolation of symptomatic cases. The timing of isolation is drawn from the onset-to-isolation distribution (see \nameref{epiparameters} section for parameterisation). All infection times that are sampled to occur after the isolation of an infector are not infected. Therefore, the isolation of a case can reduce the number of secondary cases if isolation happens prior to any of the secondary infection times. \\

@@

In this study we evaluate the feasibility of controlling outbreaks of selected influenza subtypes that have previous caused sporadic human-to-human outbreaks with pandemic potential. We include A/H5N1, A/H1N1, and A/H7N9. These have caused outbreaks of various size and severity over the past 100 years. H5N1 has caused small outbreaks ($<$ 10 cases) of human-to-human transmission \citep{yangDetectingHumanhumanTransmission2007a, aditamaAvianInfluenzaH5N12012a}. H1N1 is now an endemic seasonal influenza with substantial population immunity, but a novel strain, H1N1pdm09, emerged in 2009 and caused a pandemic outbreak \citep{fraserPandemicPotentialStrain2009, lesslerOutbreak2009Pandemic2009} (ref on history of H1N1pdm). \\

@@

>>>>

Infectious disease outbreaks pose an everpresent public health threat. The COVID-19 pandemic caused over 7 million deaths \citep{whocovid-19dashboardCOVID19DeathsWHO}, while the 1918 influenza pandemic caused an estimated $\sim$50 million deaths \citep{johnsonUpdatingAccountsGlobal2002}. Less transmissible pathogens, such as Ebola, can cause sustained outbreaks leading to the excess death of thousands \citep{whoebolaresponseteamEbolaVirusDisease2014}. Thus, timely and accurate response is critical early in an outbreak when the pandemic or epidemic potential is still uncertain \citep{kucharskiControllingMinorOutbreaks2024}. An effective response requires understanding the efficacy and feasibility of intervention strategies across pathogens with varying epidemiological characteristics \citep{fraserFactorsThatMake2004}. \\

The likelihood of an outbreak being controlled, in other words going locally-extinct or -eliminated, is in part based on the epidemiological characteristics of the pathogen, and in part on the inventions implemented, both pharmaceutical and non-pharmaceutical. The primary epidemiological determinant for the spread of an infectious disease is the reproduction number ($R$), defined as the average number of secondary cases  per primary case. The higher the reproduction number the less likely a disease, without intervention, is to cease transmission in a population. If a pathogen is subcritical (i.e. $R < 1$) it will eventually stop spreading but can still cause transient outbreaks \citep{farringtonDistributionTimeExtinction1999}. Individual-level heterogeneity in transmission also influences the possibility of a pathogen going extinct from stochasticity in the transmission process \citep{lloyd-smithSuperspreadingEffectIndividual2005}. Higher individual-level variation for a given reproduction leads to, on average, more outbreaks being controlled (however a small number of outbreaks may become large due to superspreading). In addition, the likelihood of containment is reduced --- assuming $R > 1$ --- if there are multiple introductions into a populations \citep{kucharskiEarlyDynamicsTransmission2020}. \\

A range of interventions can be deployed to reduce or prevent the transmission of infectious diseases, broadly classified into pharmaceutical and non-pharmaceutical interventions (NPIs). Early in an outbreak --- especially in the case of a novel pathogen --- NPIs are often the first line of defence, as pharmaceutical options may not yet be available. These interventions vary in scope and specificity, from targeted measures such as contact tracing and isolation, to broader actions like school closures, bans on gatherings, and regional or national lockdowns (Huag et al., 2020). The success of such interventions in containing an outbreak before it becomes sustained and widespread depends on several factors, including the type of intervention, the delay before implementation, and the intervention’s effectiveness (Longini et al., 2005). For highly transmissible pathogens (e.g. $R>3$) with short generation times (e.g. $<3$ days), swift and effective interventions are required to achieve containment. Pathogens that have a large proportion of onward transmission prior to symptom onset (latent period shorter than incubation period) or have asymptomatic (subclinical) cases are less easily controlled \citep{fraserFactorsThatMake2004}. \\

Contact tracing, the process of finding the contacts of infected individuals is an often used NPI to try and curtail the growth of an epidemic. It can involve forward contact tracing, whereby infected individuals are asked who they have in contact with so their contacts can be followed-up and isolated or quarantined if needed; or backward contact tracing whereby infected individuals are asked who they have been in contact with prior to being infected to identify the source of infection and potentially forward trace from the source. Contact tracing and isolation is a targeted approach, like ring vaccination, that looks to prevent the onward transmission of an infectious disease by tackling infected individuals and their contacts \citep{kucharskiEffectivenessRingVaccination2016, ferrettiQuantifyingSARSCoV2Transmission2020, whittakerQuantifyingImpactBroadly2024} (Keeling et al). Isolation or vaccination of traced individuals has proved widely effective in controlling different outbreaks with a variety of pathogens differing in their aetiological and epidemiological characteristics \citep{foegeSELECTIVEEPIDEMIOLOGICCONTROL1971, bellPublicHealthInterventions2004}. However, it's important to remember that it's not a panacea for all outbreak scenarios. In epidemics or pandemics where the proportion of cases or contacts missed is high, either because surveillance is patchy or because the number of contacts exceeds the capacity of the system, the efficacy of contact tracing to contain an outbreak diminishes, or at worst, fails completely \citep{dhillonWhenContactTracing2018, hellewellFeasibilityControllingCOVID192020}. \\

It has become widely accepted that influenza viruses cannot be controlled with contact tracing and isolation because the rate at which infected individuals become infectious and show symptoms is faster than the latency in the process of contacting contacts (Fraser et al., 2004; Klinkenberg et al., 2006). However, with the recent human infections of avian influenza (A/H5N1), mostly among poultry and cattle farmers (Garg et al., 2025) and the longer estimates of incubation period and serial interval (SI) from historical H5N1 data \citep{Ward2024.12.11.24318702} it may be that contact tracing and isolation is able to control influenza. The (antigenic) novelty of a pathogen introduced into a human population is a leading risk factor for causing a pandemic. Thus, influenza subtypes that have negligible levels of existing immunity in the population are probable epidemic and pandemic candidates. This is the case for H5N1 and H7N9 \citep{tannerPandemicPotentialAvian2015} and was the case when H1N1 caused an outbreak in 2009 \citep{fraserPandemicPotentialStrain2009}. Historically, avian influenza (H5N1 and H7N9) have had high severity (case fatality risk $\sim 35\%-55\%$), and therefore sustained transmission in a human population could have substantial mortality and morbidity implications \citep{tannerPandemicPotentialAvian2015}. \\

In this study we evaluate the effectiveness of controlling infectious disease outbreaks that exhibit influenza-like epidemiological characteristics using a mathematical model. In effect the ability to control an epidemic with contact tracing and isolation is a race between the identification and isolation of contacts and the spread of the infectious agent (Reyna-Lara et al., 2021). The dogma that influenza contagion outpaces our ability to trace and isolate in time, may not hold for all pathogen subtypes with new estimates of incubation period and serial interval (Ward et al., 2024). This study looks to define the speed and effectiveness of isolation under different response delays across epidemic scenarios to inform outbreak response; quantifying the impact of reducing the delay to isolation on outbreak containment and the utility of rapid testing.

@@

The NPI implemented in the transmission model is an isolation of symptomatic cases. The timing of isolation is drawn from the onset-to-isolation distribution (see \nameref{epiparameters} section for parameterisation). All infection times that are sampled to occur after the isolation of an infector are not infected. Therefore, the isolation of a case can reduce the number of secondary cases if isolation happens prior to any of the secondary infection times. Another way to iterpret the model is that $R$ is composed of secondary infections before isolation ($R_{pre}$) and secondary infections after an assumed isolation time ($R_{isolated}$).  \\

@@

In this study we evaluate the feasibility of controlling outbreaks of selected influenza subtypes that have previous caused sporadic human-to-human outbreaks with pandemic potential. We include A/H5N1, A/H1N1, and A/H7N9. These have caused outbreaks of various size and severity over the past 100 years. H5N1 has caused small outbreaks ($<$ 10 cases) of human-to-human transmission \citep{yangDetectingHumanhumanTransmission2007a, aditamaAvianInfluenzaH5N12012a}. H1N1 is now an endemic seasonal influenza with substantial population immunity, but a novel strain, H1N1pdm09, emerged in 2009 and caused a pandemic outbreak \citep{fraserPandemicPotentialStrain2009, lesslerOutbreak2009Pandemic2009}. \\

@@

\item Our branching process model used a simple forward contact-tracing mechanism that reduces the number of secondary contacts based on the timing of isolation of symptomatic cases. This contact tracing stragety is minimal cost and effort. In scenarios where outbreaks are uncontrolled a more comprehensive and costly contact tracing system where infectors and contacts of infectors (i.e. backward and forward contact tracing) are followed up, which has been shown to be more effective at outbreak containment (Klinkenberg et al., 2006).
\item Contact tracing and isolation complementing in combination with other quickly available measures for pandemic potential influenza strains, such as rapid testing and antivirals \citep{haydenPerspectivesAntiviralUse2001}.
